Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-10-17 response_to_reviewers.Rmd

## Major comments from Reviewer 2

> While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.

>Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.

> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the "big ideas" in sports analytics given the focus on this topic in many sports.  See, for example, Mark Glickman's recent presentation on this at the 2022 Joint Statistical Meetings.  I don't think that Hvattum (2019) is sufficient to cover the area of player contribution.

@@

## Minor comments from Reviewer 1

> In the second sentence, baseball is introduced as a "naturally discrete" sport, but there are certainly many things about baseball which are not discrete, as talked about later in the article. It would be helpful to clarify what is meant by discrete in this setting.

@@

We have added a paragraph clarify the distinction between "discrete" and "continuous" sports. In order to maintain the flow of the narrative, the paragraph does not appear here, but rather at the beginning of Section 2.

> Also, later on in the first paragraph, it would be useful to have a citation or two about the changes being brought about in baseball due to analytics.

We believe that the last sentence of the first paragraph does exactly that.

> In Section 2.1, Figure 1 (George Lindsey's calculations) is introduced before the reader even knows who George Lindsey is.

@@

We think this is mainly due to LaTeX floating the figure suboptimally, and this can be corrected in the page proofs. We have moved the sentence introducing Figure 1 and Table 1 until after the mention of Lindsey's work in the text to mitigate this problem.

> Furthermore, the expected run matrix is displayed "in its more familiar form", but I would venture for most readers of this article, no form would be familiar.

@@

We have replaced "more familiar" with "most common".

> I think the caption does explain the matrix nicely, although it might be useful to point out from which column in Lindsey's table the matrix is derived.

@@

Indeed, we note which column corresponds to the expected run matrix in the caption.

> In Figure 2, I found it quite surprising that the expected point value was so negative at the 95 yard line. Were safeties very common in this era of American football? This might be worth mentioning/discussing.

We have added a sentence to address this. The expected value was unrealistic because the model is linear, which is a flaw of this approach.

@@

> There is a large body of literature on Bayesian state-space models in sports outside of the Glickman and Stern paper, mainly in soccer/association football. See for example Koopman and Lit (2015) and the various references in its introduction.

Thank you for this suggestion. We have added a reference to Koopman's paper.

> As mentioned earlier, Section 5 seems sparse on details, and furthermore it seems odd to choose chess as the case study when it has barely been mentioned previously in the article. I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level. Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.

We have added a table summarizing the sport-specific packages that are available on CRAN.

> Figure 6 doesn't add much to the text; I don't think the reader better understands sports analytics by visualizing many team colors. Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e. show how that package might be used in a real-world setting.

@@

XX

> For Section 6, there is also now a "Big Data Derby" in horse racing.

@@

We thank the reviewer for this suggestion. We have added a reference to it.

> All in all, I think the article explains the ideas well and covers a lot of ground, but more care should be taken towards the organization and what to include as figures. The last half of the paper has a very different pace than the first half. I think a common theme in the paper is the rise in tracking data and how that has impacted sports–perhaps that idea could be better unified and stated in a specific section of the article, or at the very least noted in the conclusion.

XX

> Page 4, line 56: Needs to be some sort of separation between the "absorbing states" and "touchdown"

@@

## Minor comments from Reviewer 2

> Pages 3,4, 5: The use of historical tables and data is powerful.  However, I worry about the quality of the images particularly Figure 2.

We share this concern. We are also uncertain about any licensing restrictions. We have left the images as is for now, and will wait direction from Wiley.

> Page 6, line 8:  "the effect may be overstated" is somewhat ambiguous.  Is it the

@@

XX Quang or Greg to respond

> Page 6, line 53: Is the "!" warranted here for conditioning on a 9-dimensional vector?

@@

XX We could consider this.

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

XX Yeah

> 12, 37: Theming *teamcolors* is a great *R* package, not sure it is part of the "big ideas" in sports analytics.  A single line in the first paragraph of this section would seem to be better.

@@

We have added examples from the `SportyR` package, which draws playing surfaces, to help flesh out this subsection.

> 13, 35: Add a citation for Big Data Cup, https://www.stathletes.com/big-data-cup/, for SABR Diamond Dollars Case Competition, https://sabr.org/analytics/case, Big Data Derby https://www.kaggle.com/competitions/big-data-derby-2022, and maybe more.

Thank you for these suggestions. We have added them.

> 13, 43-53: don't think there is a need for a bulleted list here.

We have removed the bulleted lists.

### 2022-10-17 sports_analytics.Rmd
address #35
#32 and #36: elo with greg chess example

The task of estimating expected points of possession in football goes back to @Carter1971operations, who estimate the expected points for 1st and 10 plays in the NFL, given any yard line on the football field. 
Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
Figure \ref{fig:carter} shows a reproduction of @Carter1971operations's estimates. 
As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.

@@

## Elo ratings

@@

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in a game recently won by one of the authors (downloaded via the \pkg{chessR} package).

>>>>

The task of estimating expected points of possession in football goes back to @Carter1971operations, who estimate the expected points for 1st and 10 plays in the NFL, given any yard line on the football field. 
Due to limitations regarding the amount of data collected, the authors divide football field into 10-yard buckets, centered at their midpoints (e.g. 5, 15, 25, 35, etc.), before averaging the value of the next scoring instance across the field to obtain the expected points.
Figure \ref{fig:carter} shows a reproduction of @Carter1971operations's estimates. 
As expected, the estimated expected points increases monotonically as the teams gets closer to the endzone.
One limitation of this approach is the linearity assumption, which results in a high negative expected point value when the offensive team is 95 yards away from the opponent's goal line.

@@

## Elo ratings {#sec:elo}

@@

```{r}
library(chessR)
us <- c("fatgreggy")
greg <- get_raw_chessdotcom(
  usernames = us,
  year_month = 202206
)
# greg: black
```

To illustrate, we consider a chess game played on June 1, 2022 on [Chess.com](https://www.chess.com/) by one of the authors, with data obtained from the \pkg{chessR} package [@R-chessR]. 
Prior to the game, the author was rated `r greg[6, "BlackElo"]`, whereas `r greg[6, "WhiteElo"]` was the opponent's rating.
Since both ratings are below 2400, we apply a development coefficient of $K=20$ for this example.
The probability of author ($a$) defeating their opponent ($b$) is
<!-- don't know how to include `r` inside latex -->
$$
\Pi_{ab} = \frac{1}{1 + 20^{(1683 - 1732)/400}} = 0.591 \,.
$$
The author that played this game ended up with a victory, which means the associated score is $S_a = 1$.
The post-game Elo rating for this author is
$$
R^*_a = 1732 + 20(1 - 0.591) = 1740 \,.
$$

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package).
